White Dwarf Mergers on Adaptive Meshes\\ III. Mergers

### 2016-07-20 papers/wdmerger_III/paper.tex
Clarify some wording for hybrid advection

way that we actually discretize them for use in the code (by use of the product rule
to rearrange the terms with derivatives in them):

@@

    - \left(\frac{\partial(p (x/R))}{\partial x} + \frac{\partial(p (y/R))}{\partial y}\right) \notag \\
    &\hspace{0.175in} + \frac{1}{\rho R^3}\left(\ell_z^2 + p\rho R^2\right) \label{eq:radial-momentum-rearranged}\\

@@

It is clear that the second term on the right-hand-side in each case can be swept up

@@

is the last term on the right-hand-side of the radial momentum equation, which
exists as a cell-centered source term (which we deal with using a predictor-corrector
method similar to how we implement external forcing like gravity and rotation; see
\cite{castro} for details). It is important to observe that in this formulation
the radial momentum equation is \textit{not} a conserved quantity: it should and
will in general change with time over the course of the simulation. The above form,
absent of source terms, makes this clear. The angular momentum term in 
\autoref{eq:angular-momentum} can be thought of physically as expressing the fact
that a particle with some initial angular momentum will have its radial momentum
increase with time as it moves away from the origin, and can be thought of geometrically
as expressing the curved nature of the underlying coordinates for the radial momentum
\citep{motl:2002}. Note that while the version in \autoref{eq:radial-momentum-rearranged}
is more convenient for our hydrodynamics methodology, the pressure-gradient terms
in the first version are the ones appropriate for a radial field and so no physical meaning
should be ascribed to the pressure term at the end of \autoref{eq:radial-momentum-rearranged}.

@@

To implement this method in \castro, we desired an approach that was flexible
and would leave as little imprint on the code structure as possible.
To this end, what we have done is to add three new ``hybrid'' momentum state
variables, corresponding to the radial momentum, angular momentum, and the linear
momentum component that is perpendicular to both. Whenever we update the normal
momentum state variables with hydrodynamics fluxes, we add the analogous fluxes
to the hybrid momenta -- that is, using the edge state values determined by the
final multi-dimensional Riemann solve, we construct the value of the hybrid
momenta on that zone edge, and then allow it to be transported as usual by the
advective velocity $\mathbf{u}$. When we update the normal momenta with an
external forcing $\mathbf{F}$ such as gravity or rotation, with $x$ and $y$
components $F_x$ and $F_y$, we apply the update to the hybrid momenta as well:
\begin{align}
  \left.\frac{\partial(s_R)}{\partial t}\right|_{F} &= -F_x \frac{x}{R} - F_y\frac{y}{R} \label{eq:force-radial}\\
  \left.\frac{\partial(\ell_z)}{\partial t}\right|_{F} &= F_x\, y - F_y\, x.\label{eq:force-angular}
  \end{align}
Then, when we are in a phase of the evolution when we want to conserve angular
momentum, we make one additional change: at the end of a (single-level) advance,
we recompute the normal momenta so that they are fully consistent with the hybrid
momenta. This is all that is necessary to have a calculation that conserves angular
momentum over the course of the simulation, without changing any of the
core infrastructure for the hydrodynamics update. For example, when we compute
the primitive variables from the conservative variables, we will still be
getting them from the normal momenta, but these normal momenta are always consistent
with the conserved angular momentum, so we get the same result as if we had done
a much more involved update where we explicitly computed the primitive variables
using the hybrid momenta. There is of course no unique choice for how to do this.
For example, we could take the approach of explicitly re-calculating the normal
momenta after any change has been made to the state. This would add significant code
complexity without a clear benefit. The most significant difference would be
that when evaluating the new-time value of source terms like rotation, the velocity
field used is slightly different than it would otherwise be; but this is a high-order
effect that does not change the core angular momentum conservation property. An
analysis of the different mechanisms for implementing this update may be an interesting
avenue for future research --- including various aspects like how to implement the
source terms in the radial momentum equation. Of course, the answer will likely
be problem-dependent. Indeed, we already faces this situation in general with the
gravitational and rotational source terms, even independent of the hybrid momentum
discussion --- the velocity field seen by the new-time rotation source terms depends
on whether we apply the gravitational force before or after the rotation force.

>>>>

way that illustrates how the update is done in the code:

@@

    - \frac{1}{R}\left(x \frac{\partial p}{\partial x} + y \frac{\partial p}{\partial y}\right)
    + \frac{\ell_z^2}{\rho R^3} \label{eq:radial-momentum-rearranged}\\

@@

The first term is the standard advective flux term for any variable that advects with
the flow, and the state quantity inside this advective term can be derived on cell edges
using the primitive hydrodynamic variables after a Riemann solver has been performed.
The second term on the right-hand-side of \autoref{eq:angular-momentum-rearranged} can be swept up

@@

are the last two terms on the right-hand-side of \autoref{eq:radial-momentum-rearranged}.
While maintaining second-order accuracy in time, the last term can be treated as a
cell-centered source term that we deal with using a predictor-corrector
method (similar to how we implement external forcing like gravity and rotation; see
\cite{castro} for details). The pressure gradients in the second-to-last term can be
constructed using the Riemann-solved edge states, and they are multiplied by zone
coordinates that are cell-centered. In principle this is second-order accurate in time
because the edge states are too; however, as a drawback, the pressure is unaware of the
source terms that are applied in the corrector step after a hydrodynamics update. The
benefit of this approach is that we do not need to construct a general cell-centered
gradient of the pressure using cell-centered state data, nor do we need to perform the
equation of state call that entails.

It is important to observe that in this formulation
the radial momentum equation is \textit{not} a conserved quantity: it should and
will in general change with time over the course of the simulation. The above form,
absent of source terms, makes this clear. The angular momentum term in 
\autoref{eq:angular-momentum} can be thought of physically as expressing the fact
that a particle with some initial angular momentum will have its radial momentum
increase with time as it moves away from the origin, and can be thought of geometrically
as expressing the curved nature of the underlying coordinates for the radial momentum
\citep{motl:2002}. Note that while the version in \autoref{eq:radial-momentum-rearranged}
is more convenient for our hydrodynamics methodology, the pressure-gradient terms
in the first version are the ones appropriate for a radial field and so no physical meaning
should be ascribed to the pressure term at the end of \autoref{eq:radial-momentum-rearranged}.

@@

To implement this method in \castro, we desired an approach that was flexible
and would leave as little imprint on the code structure as possible.
To this end, what we have done is to add three new ``hybrid'' momentum state
variables, corresponding to the radial momentum, angular momentum, and the linear
momentum component that is perpendicular to both. Whenever we update the normal
momentum state variables with hydrodynamics fluxes, we add the analogous fluxes
to the hybrid momenta -- that is, using the edge state values determined by the
final multi-dimensional Riemann solve, we construct the value of the hybrid
momenta on that zone edge, and then allow it to be transported as usual by the
advective velocity $\mathbf{u}$. When we update the normal momenta with an
external forcing $\mathbf{F}$ such as gravity or rotation, with $x$ and $y$
components $F_x$ and $F_y$, we apply the update to the hybrid momenta as well:
\begin{align}
  \left.\frac{\partial(s_R)}{\partial t}\right|_{F} &= -F_x \frac{x}{R} - F_y\frac{y}{R} \label{eq:force-radial}\\
  \left.\frac{\partial(\ell_z)}{\partial t}\right|_{F} &= F_x\, y - F_y\, x.\label{eq:force-angular}
  \end{align}
Then, when we are in a phase of the evolution when we want to conserve angular
momentum, we make one additional change: at the end of a (single-level) advance,
we recompute the normal momenta so that they are fully consistent with the hybrid
momenta. This is all that is necessary to have a calculation that conserves angular
momentum over the course of the simulation, without changing any of the
core infrastructure for the hydrodynamics update. For example, when we compute
the primitive variables from the conservative variables, we will still be
getting them from the normal momenta, but these normal momenta are always consistent
with the conserved angular momentum, so we get the same result as if we had done
a much more involved update where we explicitly computed the primitive variables
using the hybrid momenta. There is of course no unique choice for how to do this.
For example, we could take the approach of explicitly re-calculating the normal
momenta after any change has been made to the state. This would add significant code
complexity without a clear benefit. The most significant difference would be
that when evaluating the new-time value of source terms like rotation, the velocity
field used is slightly different than it would otherwise be; but this is a high-order
effect that does not change the core angular momentum conservation property. An
analysis of the different mechanisms for implementing this update may be an interesting
avenue for future research --- including various aspects like how to implement the
source terms in the radial momentum equation. Of course, the answer will likely
be problem-dependent. Indeed, we already face this situation in general with the
gravitational and rotational source terms, even independent of the hybrid momentum
discussion --- the velocity field seen by the new-time rotation source terms depends
on whether we apply the gravitational force before or after the rotation force.
